Joint maximum-likelihood of phylogeny and ancestral states is not consistent

### 2018-11-06 appendix.tex
going through comments

%EM This is very minor compared to the below comments, so address it second in a separate commit, please, if you wish to address it. Curly braces are really for unordered sets, right? I suggest round parensI suggest round parens.
%das: TODO
For a fixed topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization $\ancestralStateColumn_i$.
In this case we cannot use the same complement trick as before: the probability of observing an ancestral state split conditional on a site split is not invariant to taking its complement.
We thus define an ``ancestral state split'' $\ancestralSplit$ for an internal node $\ancestralStateColumn_i$ to be the set of internal nodes labeled with $1$ if the $\nSiteRows$th tip is not labeled with $1$, and as the set of internal nodes labeled with $1$ in $\overline{\ancestralStateColumn}_i$ if the $\nSiteRows$th tip is labeled with $1$.
We emphasize that the ancestral state split complementing procedure depends on tip states, not ancestral states: both site splits and ancestral state splits are defined by whether the $\nSiteRows$th element of $\alignmentColumn_i$ is labeled as $1$.

@@

The likelihood in \eqref{eq:profile_likelihood} can be written as
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t) \nonumber \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
for $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ and some $\xi_{\siteSplit_j}\in\ancestralSplitPartition_j(\tau, t)$ with $1 \le j \le q$ where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.

@@

%das: @EM I think this is actually the same Wald reference that Felsenstein tried to use.
%It works here for a fixed topology, and didn't work for him or your reference because I think the upper-semicontinuous condition doesn't hold for trees in general.
%Also I don't think the compactness condition holds---tree space isn't compact is it?
%However we now have a potential snag in that we are doing what Yang (1994) in his intro says is bad, i.e., doing ML for fixed, separate topologies and comparing them.
%We still have the behavior in the plots and the proofs, since they all assumed a fixed topology, and I think everything's fine with the arguments about Fels vs InvFels, but maybe you can let me know what you think?
%EM @das I think that ^ is all resolved, right?
%das: I think to be fair I should add that we don't actually show a real inconsistency in topology because we haven't shown this functional convergence for it, only for branch lengths with a fixed topology. I added a paragraph hedging a bit. Let me know what you think.
For a fixed $\tau$, we show that $\hat{t}_\nCols \rightarrow \hat{t}$ for
\[
\hat{t}_\nCols = \arg\max_{t\in\mathcal{T}} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
\]
and
\[
\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
\]
Using the notation in Section 5.2.1 in \citet{van1998asymptotic}, we let
\[
m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)
\]
so that
\[
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) = \frac{1}{n} \sum_{i=1}^\nCols m_t(\alignmentColumn_i)
\]
and
\[
\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = E[m_t].
\]
To show $\hat{t}_\nCols \rightarrow \hat{t}$, we use Wald's consistency proof \citep[p. 48, Theorem 5.14 of ][]{van1998asymptotic}, which requires four conditions.
The first is that $\mathcal{T}$ is compact, which is obviously true.
The second is that
\[
E\left[\sup_{t\in\mathcal{T}} m_t\right] < \infty,
\]
and, since $m_t(\alignmentColumn)$ is nonpositive for all $t$ and $\alignmentColumn$, this property holds.
The remaining conditions are on the maps
\[
\alignmentColumn \mapsto \sup_t m_t(\alignmentColumn)
\]
and
\[
t \mapsto m_t(\alignmentColumn).
\]
We need the first map to be measurable, which is evident since the domain $\alphabet^\nSiteRows$ of the mapping is a finite set, and so all subsets of the domain are also finite and thus measurable.
Finally, we must have the the second mapping be upper-semicontinuous for almost all $\alignmentColumn$.
For a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
If we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we maximize over ancestral state splits we obtain a continuous function since the maximum over continuous functions is also continuous.
This ensures the upper-semicontinuous property of this mapping, and shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$.

@@

\begin{proof}
For a fixed, known $\boldsymbol\xi$, there exists a closed form solution to $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ solving
\[
\hat{t}_{\boldsymbol\xi} = \arg\max_t \ell_{\tau^*, t^*}(\tau_1, t; \boldsymbol\xi).
\]
We show in this case that the log-likelihood $\ell$ attains a unique maximum at $\hat{t}_{\boldsymbol\xi}$.
For fixed $\boldsymbol\xi$, the log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
%EM The previous version switched between only looking at x_1 and the general case. Here I tried to make that more flexible and explicit.
\begin{align*}
\ell_{\tau^*,t^*}(\tau^*, t, \boldsymbol\xi) &= \sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,x_1}(x_1)+\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,y_1}(y_1)+\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,x_2}(x_2)\\
&\qquad + \sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,y_2}(y_2)+\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,w}(w).
\end{align*}
Due to this additive form, all off-diagonal terms of the Hessian for this function are zero, so we show that the diagonal terms are nonpositive.
For example, if we focus on the variable $x_1$, holding other variables constant, then
\[
\ell(x_1) \sim \sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,x_1}(x_1).
\]
where $\sim$ denotes equality up to an additive constant.
Doing calculation as in Figure~\ref{fig:example_likelihoods}, each functional form, suppressing constants with respect to the focal variable (here $x_1$) and the initial $1/32$ constant, takes the form
\[
h_{j,x_1}(x_1) = (1+x_1)^{e_j}(1-x_1)^{1-e_j}
\]
for $e_j\in\{0,1\}$, which, simplifying, results in
\begin{align}
\ell(x_1) &\sim \left(\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}(1-e_j)\right) \log (1-x_1)\\
                  &= \left(\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) \log (1-x_1), \label{eq:closed-form-x}
\end{align}
which has second derivative
\[
\ell''(x_1) = -\left(\frac{\sum_j p_{\siteSplit_j}e_j}{(1+x_1)^2}+\frac{1-\sum_j p_{\siteSplit_j}e_j}{(1-x_1)^2}\right).
\]
As $x_1\in(0,1]$, we need only $0 \le \sum_j p_{\siteSplit_j}e_j \le 1$ to imply the diagonal terms of the Hessian are nonpositive.
Since $p_{\siteSplit_j}$ are probabilities, then $\sum_j p_{\siteSplit_j} = 1$.
As $e_j\in\{0,1\}$ by definition, this implies $0 \le \sum_j p_{\siteSplit_j}e_j \le 1$ and $\ell''(x_1) \le 0$.
Applying similar arguments to the other variables, the Hessian for the log-likelihood has nonpositive diagonal terms and off-diagonal terms equal to zero, and $\hat{t}$ uniquely maximizes $\ell$.

@@

Next we show that solutions of this form never obtain $\hat{t} = t^*$ except on a set of measure zero.
Given Table~\ref{tab:gen-sitepatprob}, all solutions $\hat{x}_1$ have the form
\[
\hat{x}_1(x^*,y^*) = a_{x_1,0} + a_{x_1,1}(x^*)^2 + a_{x_1,2}(y^*)^2 + a_{x_1,3} x^*(y^*)^2 + a_{x_1,4}(x^*)^2(y^*)^2,
\]
where $a_{x_1,k}$ are constants independent of $x^*$ and $y^*$---in fact, $a_{x_1,k}$ takes values in the set $\{i/8 : i=-4,-3,\ldots,7,8\}$.
The true branch fidelity for $x_1$ is $x^*$, and so we have consistency when
\[
f_{y^*}(x^*) = \hat{x}_1(x^*,y^*) - x^*
\]
is zero.
As the number of zeros of $f_{y^*}(x^*)$ is finite for $0 < x^*, y^* < 1$, consistency holds only on a set of measure zero.
The same is true for $x_2$, and a similar argument for $y_1,y_2,$ and $w$ shows that estimates can only be consistent on a set of measure zero.

%EM This is much improved, thanks! It seems like we need a little more to complete the proof in the general case. We start the proof hypothesizing a fixed \xi. When do we relax that assumption?
\end{proof}

%EM I don't see this paragraph being helpful.
%EM It's a confusing read. First you say that we're going to look at examples, but then perhaps you cast doubt on if the examples exist "There may be no..."
%das: moved some things around so the example is outside of the proof
As an example solution for maximizing $x_1$, the maximal ancestral state splits for $\hat{\boldsymbol\xi}_1 = \{\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset\}$ (Table~\ref{tab:likelihoods}) yields the log-likelihood
\begin{align*}
\ell_{\tau^*,t^*}(x_1; \hat{\boldsymbol\xi}_1) &= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
&\qquad + \left(\frac{1}{2}+\frac{1}{2}(y^*)^2\right)\log(1+y_1) + \left(\frac{1}{2}-\frac{1}{2}(y^*)^2\right)\log(1-y_1)\\
&\qquad + \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_2) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_2)\\
&\qquad + \log(1+y_2) + \log(1+w) - \log 32\\
&:= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
&\qquad + C_{x_1}(x_2,y_1,y_2,w).
\end{align*}
where $C_{x_1}$ is a function of $y_1,x_2,y_2,$ and $w$ and
\[
\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j = p_{\emptyset} + p_2 + p_3 + p_{23} = \frac{1}{2} + \frac{1}{2}x^*(y^*)^2.
\]
Thus, $\hat{x}_1(x^*, y^*) = x^*(y^*)^2$, and it is possible to obtain $\hat{x}_1(x^*,y^*)=x^*$, such as when $y^*=1$, but we cannot have $\hat{x}_1(x^*,y^*)=x^*$ for all $x^*$ where $0 < x^*, y^* < 1$.

@@

The boundaries determining maximal ancestral state splits (Fig.~\ref{fig:max-anc-state}) are obtained through maximizing over 81 separate functional values (Table~\ref{tab:likelihoods}).
Referring to the proof of Theorem~\ref{thm:bl_theorem}, we see that
\[
\ell(x_1) \sim \left(\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) \log (1-x_1)
\]
and
\[
\hat{x}_1 = 2\cdot\left(\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) - 1
\]
so that the maximal value for $\ell(x_1)$ is
\[
\frac{1+\hat{x}_1}{2}\log(1+\hat{x}_1) + \frac{1-\hat{x}_1}{2}\log(1-\hat{x}_1).
\]
%EM The following sentence needs revision.... read aloud?
With similar forms for the remaining variables our likelihood given in~\eqref{eq:site_pattern_profile_likelihood_mean} has an exact, closed form maximum of
\begin{align}
\ell_{\tau^*,t^*}(\tau_1, \hat{t}; \boldsymbol\xi) &= \frac{1+\hat{x}_1}{2}\log(1+\hat{x}_1) + \frac{1-\hat{x}_1}{2}\log(1-\hat{x}_1)\nonumber\\
&\qquad + \frac{1+\hat{y}_1}{2}\log(1+\hat{y}_1) + \frac{1-\hat{y}_1}{2}\log(1-\hat{y}_1)\nonumber\\
&\qquad + \frac{1+\hat{x}_2}{2}\log(1+\hat{x}_2) + \frac{1-\hat{x}_2}{2}\log(1-\hat{x}_2)\nonumber\\
&\qquad + \frac{1+\hat{y}_2}{2}\log(1+\hat{y}_2) + \frac{1-\hat{y}_2}{2}\log(1-\hat{y}_2)\nonumber\\
&\qquad + \frac{1+\hat{w}}{2}\log(1+\hat{w}) + \frac{1-\hat{w}}{2}\log(1-\hat{w}).
\end{align}
For example, in the lower right region of Fig.~\ref{fig:max-anc-state}, the maximal ancestral state split is $\hat{\boldsymbol\xi}_1 = \{\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset\}$ so that
\[
\{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\} = \{x^*(y^*)^2,(y^*)^2,x^*(y^*)^2,1,1\}
\]
and the exact, closed form value of the likelihood is, as a function of $x^*$ and $y^*$,
\begin{align}
g_{\boldsymbol\xi_1}(x^*,y^*) &:= (1+x^*(y^*)^2)\log(1+x^*(y^*)^2) + (1-x^*(y^*)^2)\log(1-x^*(y^*)^2)\nonumber\\
&\qquad + \frac{1+(y^*)^2}{2}\log(1+(y^*)^2) + \frac{1-(y^*)^2}{2}\log(1-(y^*)^2)\nonumber\\
&\qquad + 2\log(2).
\end{align}
We obtain similar maximal values as functions of $x^*$ and $y^*$ for each ancestral state split to get $g_{\boldsymbol\xi}$ for all relevant $\boldsymbol\xi$.
The curves delineating maximal ancestral state splits (Fig.~\ref{fig:max-anc-state}) are determined by these 81 functions.

>>>>

%EM This is very minor compared to the below comments, so address it second in a separate commit, please, if you wish to address it. Curly braces are really for unordered sets, right? I suggest round parensI suggest round parens.
%das: on second thought, this is using notation like in Semple/Steel. If we change this here, I think we'd have to change almost all curly braces in the document, including the ones that use the Semple/Steel notation. I'm not necessarily opposed, but I think it might be confusing, especially because we use their notation a lot throughout. Thoughts?
For a fixed topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization $\ancestralStateColumn_i$.
In this case we cannot use the same complement trick as before: the probability of observing an ancestral state split conditional on a site split is not invariant to taking its complement.
We thus define an ``ancestral state split'' $\ancestralSplit$ for an internal node $\ancestralStateColumn_i$ to be the set of internal nodes labeled with $1$ if the $\nSiteRows$th tip is not labeled with $1$, and as the set of internal nodes labeled with $1$ in $\overline{\ancestralStateColumn}_i$ if the $\nSiteRows$th tip is labeled with $1$.
We emphasize that the ancestral state split complementing procedure depends on tip states, not ancestral states: both site splits and ancestral state splits are defined by whether the $\nSiteRows$th element of $\alignmentColumn_i$ is labeled as $1$.

@@

The likelihood~\eqref{eq:profile_likelihood} can be written as
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t) \nonumber \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
for $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ and some $\xi_{\siteSplit_j}\in\ancestralSplitPartition_j(\tau, t)$ with $1 \le j \le q$ where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.

@@

%das: @EM I think this is actually the same Wald reference that Felsenstein tried to use.
%It works here for a fixed topology, and didn't work for him or your reference because I think the upper-semicontinuous condition doesn't hold for trees in general.
%Also I don't think the compactness condition holds---tree space isn't compact is it?
%However we now have a potential snag in that we are doing what Yang (1994) in his intro says is bad, i.e., doing ML for fixed, separate topologies and comparing them.
%We still have the behavior in the plots and the proofs, since they all assumed a fixed topology, and I think everything's fine with the arguments about Fels vs InvFels, but maybe you can let me know what you think?
%EM @das I think that ^ is all resolved, right?
%das: I think to be fair I should add that we don't actually show a real inconsistency in topology because we haven't shown this functional convergence for it, only for branch lengths with a fixed topology. I added a paragraph hedging a bit. Let me know what you think.
%das: bump
For a fixed $\tau$, we show that $\hat{t}_\nCols \rightarrow \hat{t}$ for
\[
\hat{t}_\nCols = \arg\max_{t\in\mathcal{T}} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
\]
and
\[
\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
\]
Using the notation in Section 5.2.1 in \citet{van1998asymptotic}, we let
\[
m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)
\]
so that
\[
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) = \frac{1}{n} \sum_{i=1}^\nCols m_t(\alignmentColumn_i)
\]
and
\[
\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = E[m_t].
\]
To show $\hat{t}_\nCols \rightarrow \hat{t}$, we use Wald's consistency proof \citep[p. 48, Theorem 5.14 of ][]{van1998asymptotic}, which requires four conditions.
The first is that $\mathcal{T}$ is compact, which is obviously true.
The second is that
\[
E\left[\sup_{t\in\mathcal{T}} m_t\right] < \infty,
\]
and, since $m_t(\alignmentColumn)$ is nonpositive for all $t$ and $\alignmentColumn$, this property holds.
The remaining conditions are on the maps
\[
\alignmentColumn \mapsto \sup_t m_t(\alignmentColumn)
\]
and
\[
t \mapsto m_t(\alignmentColumn).
\]
We need the first map to be measurable, which is evident since the domain $\alphabet^\nSiteRows$ of the mapping is a finite set, and so all subsets of the domain are also finite and thus measurable.
Finally, we must have the the second mapping be upper-semicontinuous for almost all $\alignmentColumn$.
For a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
If we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we maximize over ancestral state splits we obtain a continuous function since the maximum over continuous functions is also continuous.
This ensures the upper-semicontinuous property of this mapping, and shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$.

@@

\begin{proof}
For a fixed, known $\boldsymbol\xi$, there exists a closed form solution to $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ solving
\[
\hat{t}_{\boldsymbol\xi} = \arg\max_t \ell_{\tau^*, t^*}(\tau_1, t; \boldsymbol\xi).
\]
We show in this case that the log-likelihood $\ell$ attains a unique maximum at $\hat{t}_{\boldsymbol\xi}$.
For fixed $\boldsymbol\xi$, the log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
%EM The previous version switched between only looking at x_1 and the general case. Here I tried to make that more flexible and explicit.
%das: looks good :+1:
\begin{align*}
\ell_{\tau^*,t^*}(\tau^*, t, \boldsymbol\xi) &= \sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,x_1}(x_1)+\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,y_1}(y_1)+\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,x_2}(x_2)\\
&\qquad + \sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,y_2}(y_2)+\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,w}(w).
\end{align*}
Due to this additive form, all off-diagonal terms of the Hessian for this function are zero, so we show that the diagonal terms are nonpositive.
For example, if we focus on the variable $x_1$, holding other variables constant, then
\[
\ell(x_1) \sim \sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,x_1}(x_1).
\]
where $\sim$ denotes equality up to an additive constant.
Doing calculation as in Figure~\ref{fig:example_likelihoods}, each functional form, suppressing constants with respect to the focal variable (here $x_1$) and the initial $1/32$ constant, takes the form
\[
h_{j,x_1}(x_1) = (1+x_1)^{e_j}(1-x_1)^{1-e_j}
\]
for $e_j\in\{0,1\}$, which, simplifying, results in
\begin{align}
\ell(x_1) &\sim \left(\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}(1-e_j)\right) \log (1-x_1)\\
                  &= \left(\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) \log (1-x_1), \label{eq:closed-form-x}
\end{align}
which has second derivative
\[
\ell''(x_1) = -\left(\frac{\sum_j p_{\siteSplit_j}e_j}{(1+x_1)^2}+\frac{1-\sum_j p_{\siteSplit_j}e_j}{(1-x_1)^2}\right).
\]
As $x_1\in(0,1]$, we need only $0 \le \sum_j p_{\siteSplit_j}e_j \le 1$ to imply the diagonal terms of the Hessian are nonpositive.
Since $p_{\siteSplit_j}$ are probabilities, then $\sum_j p_{\siteSplit_j} = 1$.
As $e_j\in\{0,1\}$ by definition, this implies $0 \le \sum_j p_{\siteSplit_j}e_j \le 1$ and $\ell''(x_1) \le 0$.
Applying similar arguments to the other variables, the Hessian for the log-likelihood has nonpositive diagonal terms and off-diagonal terms equal to zero, and $\hat{t}$ uniquely maximizes $\ell$.

@@

Next we show that solutions of this form never obtain $\hat{t} = t^*$ except on a set of measure zero.
Given Table~\ref{tab:gen-sitepatprob}, all solutions $\hat{x}_1$ have the form
\[
\hat{x}_1(x^*,y^*) = a_{x_1,0} + a_{x_1,1}(x^*)^2 + a_{x_1,2}(y^*)^2 + a_{x_1,3} x^*(y^*)^2 + a_{x_1,4}(x^*)^2(y^*)^2,
\]
where $a_{x_1,k}$ are constants independent of $x^*$ and $y^*$---in fact, $a_{x_1,k}$ takes values in the set $\{i/8 : i=-4,-3,\ldots,7,8\}$.
The true branch fidelity for $x_1$ is $x^*$, and so we have consistency when
\[
f_{y^*}(x^*) = \hat{x}_1(x^*,y^*) - x^*
\]
is zero.
As the number of zeros of $f_{y^*}(x^*)$ is finite for $0 < x^*, y^* < 1$, consistency holds only on a set of measure zero for a fixed $\boldsymbol\xi$.
Since we have inconsistency in $\hat{t}_{\boldsymbol\xi}$ for each $\boldsymbol\xi$, this implies an inconsistency when maximizing over $\boldsymbol\xi$ as it takes values on a finite set.

The same is true for $x_2$, and a similar argument for $y_1,y_2,$ and $w$ shows that estimates can only be consistent on a set of measure zero.

%EM This is much improved, thanks! It seems like we need a little more to complete the proof in the general case. We start the proof hypothesizing a fixed \xi. When do we relax that assumption?
%das: yah, maximizing over xi is straightforward since it takes values on a finite set. I changed a bit of notation so hope it's more clear now.
\end{proof}

%EM I don't see this paragraph being helpful.
%EM It's a confusing read. First you say that we're going to look at examples, but then perhaps you cast doubt on if the examples exist "There may be no..."
%das: moved some things around so the example is outside of the proof
%das: bump
As an example solution for maximizing $x_1$, the maximal ancestral state splits for $\hat{\boldsymbol\xi}_1 = \{\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset\}$ (Table~\ref{tab:likelihoods}) yields the log-likelihood
\begin{align*}
\ell_{\tau^*,t^*}(x_1; \hat{\boldsymbol\xi}_1) &= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
&\qquad + \left(\frac{1}{2}+\frac{1}{2}(y^*)^2\right)\log(1+y_1) + \left(\frac{1}{2}-\frac{1}{2}(y^*)^2\right)\log(1-y_1)\\
&\qquad + \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_2) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_2)\\
&\qquad + \log(1+y_2) + \log(1+w) - \log 32\\
&:= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
&\qquad + C_{x_1}(x_2,y_1,y_2,w).
\end{align*}
where $C_{x_1}$ is a function of $y_1,x_2,y_2,$ and $w$ and
\[
\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j = p_{\emptyset} + p_2 + p_3 + p_{23} = \frac{1}{2} + \frac{1}{2}x^*(y^*)^2.
\]
Thus, $\hat{x}_1(x^*, y^*) = x^*(y^*)^2$, and it is possible to obtain $\hat{x}_1(x^*,y^*)=x^*$, such as when $y^*=1$, but we cannot have $\hat{x}_1(x^*,y^*)=x^*$ for all $x^*$ where $0 < x^*, y^* < 1$.

@@

The boundaries determining maximal ancestral state splits (Fig.~\ref{fig:max-anc-state}) are obtained through maximizing over 81 separate functional values (Table~\ref{tab:likelihoods}).
Referring to the proof of Theorem~\ref{thm:bl_theorem}, we see that
\[
\ell(x_1) \sim \left(\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) \log (1-x_1)
\]
and
\[
\hat{x}_1 = 2\cdot\left(\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) - 1
\]
so that the maximal value for $\ell(x_1)$ is
\[
\frac{1+\hat{x}_1}{2}\log(1+\hat{x}_1) + \frac{1-\hat{x}_1}{2}\log(1-\hat{x}_1)
\]
%EM The following sentence needs revision.... read aloud?
%das: changed
with similar forms for the remaining variables.
Our likelihood~\eqref{eq:site_pattern_profile_likelihood_mean} has an exact, closed form maximum of
\begin{align}
\ell_{\tau^*,t^*}(\tau_1, \hat{t}; \boldsymbol\xi) &= \frac{1+\hat{x}_1}{2}\log(1+\hat{x}_1) + \frac{1-\hat{x}_1}{2}\log(1-\hat{x}_1)\nonumber\\
&\qquad + \frac{1+\hat{y}_1}{2}\log(1+\hat{y}_1) + \frac{1-\hat{y}_1}{2}\log(1-\hat{y}_1)\nonumber\\
&\qquad + \frac{1+\hat{x}_2}{2}\log(1+\hat{x}_2) + \frac{1-\hat{x}_2}{2}\log(1-\hat{x}_2)\nonumber\\
&\qquad + \frac{1+\hat{y}_2}{2}\log(1+\hat{y}_2) + \frac{1-\hat{y}_2}{2}\log(1-\hat{y}_2)\nonumber\\
&\qquad + \frac{1+\hat{w}}{2}\log(1+\hat{w}) + \frac{1-\hat{w}}{2}\log(1-\hat{w}).
\end{align}
For example, in the lower right region of Fig.~\ref{fig:max-anc-state}, the maximal ancestral state split is $\hat{\boldsymbol\xi}_1 = \{\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset\}$ so that
\[
\{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\} = \{x^*(y^*)^2,(y^*)^2,x^*(y^*)^2,1,1\}
\]
and the exact, closed form value of the likelihood is, as a function of $x^*$ and $y^*$,
\begin{align}
g_{\boldsymbol\xi_1}(x^*,y^*) &:= (1+x^*(y^*)^2)\log(1+x^*(y^*)^2) + (1-x^*(y^*)^2)\log(1-x^*(y^*)^2)\nonumber\\
&\qquad + \frac{1+(y^*)^2}{2}\log(1+(y^*)^2) + \frac{1-(y^*)^2}{2}\log(1-(y^*)^2)\nonumber\\
&\qquad + 2\log(2).
\end{align}
We obtain similar maximal values as functions of $x^*$ and $y^*$ for each ancestral state split to get $g_{\boldsymbol\xi}$ for all relevant $\boldsymbol\xi$.
The curves delineating maximal ancestral state splits (Fig.~\ref{fig:max-anc-state}) are determined by these 81 functions.
